---
title: "Connections between Inner Products, Vector Norms, Angles, and Sample Covariance and Correlation"
format: html
self-contained-math: true
---


This doc revisits some fundamental concepts in linear algebra and statistics: **inner products**, **vector norms**, **angles between vectors**, and how these relate to **sample covariance** and **correlation**. By the end, we’ll see how the correlation between two data vectors is simply the cosine of the angle between them. 

---

### 1. Inner Products (Scalar Products)

The **inner product** (also called the **scalar product** or **dot product**) is a way to multiply two vectors and get a single number (a scalar). For two vectors $\mathbf{u}$ and $\mathbf{v}$ in $\mathbb{R}^n$, the inner product is defined as:

$$
\mathbf{u} \cdot \mathbf{v} = \mathbf{u}^T \mathbf{v} = \sum_{i=1}^n u_i v_i
$$

This means you multiply corresponding components of the vectors and add up the results. For example, if $\mathbf{u} = [1, 2, 3]$ and $\mathbf{v} = [4, 5, 6]$, their inner product is:

$$
\mathbf{u} \cdot \mathbf{v} = (1 \cdot 4) + (2 \cdot 5) + (3 \cdot 6) = 4 + 10 + 18 = 32
$$

The inner product^[often written as $\langle u ,v \rangle$ instead of $u \cdot v$ ] has two key properties:

1. It measures to what extent the two vectors "point in the same direction" [(Detailed explanation)](https://www.3blue1brown.com/lessons/dot-products#geometric-interpretation)
2. So it is closely related to the **angle** between the two vectors.

---

### 2. Vector Norms

The **norm** of a vector $\mathbf{u}$, denoted $\|\mathbf{u}\|$, is a measure of its length. For a vector in $\mathbb{R}^n$, the norm is defined as:

$$
\|\mathbf{u}\| = \sqrt{\sum_{i=1}^n u_i^2}
$$

This is essentially the Euclidean distance from the origin to the point represented by $\mathbf{u}$. For example, if $\mathbf{u} = [3, 4]$, its norm is:

$$
\|\mathbf{u}\| = \sqrt{3^2 + 4^2} = \sqrt{9 + 16} = 5
$$

The norm is related to the inner product because:

$$
\|\mathbf{u}\| = \sqrt{\mathbf{u} \cdot \mathbf{u}}
$$

---

### 3. Angles Between Vectors

The inner product and norm are used to define the **angle** between two vectors. For vectors $\mathbf{u}$ and $\mathbf{v}$, the cosine of the angle $\theta$^[Note that maths people measure angles in *radians* (i.e., as multiples of $\pi$), not in *degrees*:  
0° is 0 radians, 90° is $\frac{\pi}{2}$ radians, 180° is $\pi$ radians, etc.] between them is given by:

$$
\cos \theta = \frac{\mathbf{u} \cdot \mathbf{v}}{\|\mathbf{u}\| \|\mathbf{v}\|}
$$

This formula connects geometry (angles) with algebra (inner products and norms). Let’s break it down:

- The numerator $\mathbf{u} \cdot \mathbf{v}$ measures how aligned the vectors are.
- The denominator $\|\mathbf{u}\| \|\mathbf{v}\|$ scales the result by the lengths of the vectors.

If $\mathbf{u}$ and $\mathbf{v}$ point in exactly the same direction, their angle is 0° and $\cos \theta = 1$. If they are perpendicular, the angle is 90° so $\cos \theta = 0$. If they point in exactly opposite directions (180° angle), $\cos \theta = -1$.

---

### 4. Sample Standard Deviation

Now, let’s connect these ideas to statistics. Suppose we have two sets of data represented as vectors $\mathbf{x} = [x_1, x_2, \dots, x_n]$ and $\mathbf{y} = [y_1, y_2, \dots, y_n]$.


Define $\bar{x}$ and $\bar{y}$ as the means of $\mathbf{x}$ and $\mathbf{y}$, respectively.
If we center the data by subtracting the means:

$$
\mathbf{x}_c = [x_1 - \bar{x}, x_2 - \bar{x}, \dots, x_n - \bar{x}]
$$

$$
\mathbf{y}_c = [y_1 - \bar{y}, y_2 - \bar{y}, \dots, y_n - \bar{y}]
$$

The standard deviations are defined as:

$$S_{X} = \sqrt{\frac{1}{n-1} \sum_{i=1}^n (x_i - \bar{x})^2} = \frac{\|\mathbf{x}_c\|}{\sqrt{n-1}}$$

and

$$S_{Y} = \sqrt{\frac{1}{n-1} \sum_{i=1}^n (y_i - \bar{y})^2} = \frac{\|\mathbf{y}_c\|}{\sqrt{n-1}}$$

We can see that they are equal to the norms of the centered data vectors, divided by the square root of their number of entries minus one.

### 5. Sample Covariance and Correlation

The **sample covariance** measures how much $\mathbf{x}$ and $\mathbf{y}$ change together:

$$
S_{XY} = \frac{1}{n-1} \sum_{i=1}^n (x_i - \bar{x})(y_i - \bar{y})
$$


The **correlation** $r_{XY}$ between $\mathbf{x}$ and $\mathbf{y}$ is a normalized version of covariance, scaled to lie between -1 and 1:

$$
r_{XY} = \frac{S_{XY}}{S_{X} S_{Y}}
$$

where $S_{X}$ and $S_{Y}$ are the standard deviations of $\mathbf{x}$ and $\mathbf{y}$.

---

### 6. Correlation as the Cosine of the Angle

Here’s the key insight: **The correlation $r_{XY}$ is exactly the cosine of the angle between the centered versions of $\mathbf{x}$ and $\mathbf{y}$.**

The correlation $r_{XY}$ can be rewritten as:

$$
r_{XY} = \frac{S_{XY}}{S_{X} S_{Y}} =  \frac{\sum_{i=1}^n (x_i - \bar{x})(y_i - \bar{y})}{\sqrt{\sum_{i=1}^n (x_i - \bar{x})^2}\sqrt{\sum_{i=1}^n (y_i - \bar{y})^2}} = \frac{\mathbf{x}_c \cdot \mathbf{y}_c}{\|\mathbf{x}_c\| \|\mathbf{y}_c\|},
$$

with $\mathbf{x}_c = \mathbf{x} - \bar x$, $\mathbf{y}_c = \mathbf{y} - \bar y$.

This is exactly the formula for $\cos \theta\,$! So:

$$
r_{XY} = \cos \theta
$$

The correlation is negative if the data vectors point in (roughly) opposite directions (their angle is between 180° and 90°, with a maximum of -1 if they are exactly opposite with an angle of 180°),
zero if they are orthogonal (their angle is 90°), and positive if the point in similar directions (their angle is betwen 90° and 0°, with a maximum of 1 if they are exactly parallel with an angle of 0°).

---

### Summary

1. The **inner product** measures alignment between vectors.
2. The **norm** measures the length of a vector.
3. The **angle** between vectors is determined by their inner product and norms.
4. The **standard deviation** of a data vector is simply the **norm of the centered data vector**, divided by the square root of their dimension.
5. **Sample covariance** measures how two data vectors vary together and is simply the **inner product of the centered data vectors** divided by their dimension.
6. **Correlation** is the cosine of the angle between the centered data vectors: their inner product divided by the product of their norms.

By understanding these relationships, you can see how linear algebra (vector norms and products), geometry (angles) and statistics (correlation) are deeply connected. A lot of the maths we need for data analysis is just applied linear algebra, so it's well worth getting a good intuitive grasp of core concepts -- I strongly recommend 3blue1brown's [Linear Algebra](https://www.3blue1brown.com/topics/linear-algebra) videos for that.

---

## Examples

Let’s work through numerical examples for this with **3 data points in 2 dimensions** to illustrate **negative**, **zero**, and **positive correlations**. We’ll use simple numbers to make the calculations easy to follow.
To do this, we essentially change our perspective on the data:  
instead of looking at the *row space* of the data matrix, that defines 3 observations/points in 2 dimensions, to see what kind of pattern these points make,  
we look at the *column space* of the data matrix, that defines 2 data vectors with 3 entries each, and compare the directions in which these vectors are pointing. 

### Setup

We’ll define two vectors $\mathbf{x}$ and $\mathbf{y}$ with 3 data points each:

$$
\mathbf{x} = [x_1, x_2, x_3], \quad \mathbf{y} = [y_1, y_2, y_3]
$$

We’ll compute:

1. The **centered vectors** $\mathbf{x_c}$ and $\mathbf{y_c}$ (by subtracting their means).
2. The **inner product** $\mathbf{x_c} \cdot \mathbf{y_c}$.
3. The **norms** $\|\mathbf{x_c}\|$ and $\|\mathbf{y_c}\|$.
4. The **correlation** $r_{XY} \frac{\mathbf{x_c} \cdot \mathbf{y_c}}{\|\mathbf{x_c}\| \|\mathbf{y_c}\|}$.
5. The **angle** $\theta = \cos^{-1}(r)$.

In the figures, we'll show

- a data table with $\mathbf{x}$, $\mathbf{y}$ and $\mathbf{x_c}$, $\mathbf{y_c}$,
- a scatter plots with three datapoints $(x_i,y_i)$, $(x_j,y_j)$, $(x_k,y_k)$,
- a 3D-plot of the corresponding $\mathbf{x_c}$ and $\mathbf{y_c}$ vectors over axes $i, j, k$.

Let’s go through each case!


```{r}
#| code-fold: true
#| code-summary: "Code for figures below"

library(plot3D)  # For 3D arrows and lines
library(patchwork) # For arranging plots side by side
library(ggplot2) # nice graphics
library(gt) #nice tables

plot_vectors <- function(data_matrix, 
                         ij = TRUE, ik = TRUE, jk = TRUE, # plot 2d-planes as visual aid?
                         theta = 50, phi = 30)  { # set viewing angles for 3d plot
  # Extract x and y vectors
  x <- data_matrix[, 1]
  y <- data_matrix[, 2]
  
  # Center the data
  x_centered <- x - mean(x)
  y_centered <- y - mean(y)
  d_centered <- cbind(x_centered, y_centered)
  
  # Create a data frame for the 2D scatterplot
  df <- data.frame(
    x = x,
    y = y,
    x_c = x_centered,
    y_c = y_centered,
    label = c("i", "j", "k")
  )
  rownames(df) <- df$label
  
  # 2D Scatterplot
  p1 <- ggplot(df, aes(x = x, y = y)) + 
    geom_point(shape = 1, size = 6, fill = "lightgrey") + 
    geom_text(label = c("i", "j", "k")) + 
    xlim(scales::expand_range(range(df$x), mul = 0.1)) +
    ylim(scales::expand_range(range(df$y), mul = 0.1)) +
    labs(x = "x", y = "y", title = "Data Points &\n Centered Data Vectors") +
    theme_minimal() + theme(axis.line.x = element_line(color = "red"),
                            axis.line.y = element_line(color = "blue"),
                            panel.grid = element_blank())
  
  p2 <- wrap_table(gt(df, rowname_col = "label"), 
                   panel = "full", space = "fixed")
  
  
  par(mar = c(0, 0, 0, 0), xaxs = "i", yaxs = "i")
  p <- 
    p2 +
    p1 + 
    # 3D Plot of Centered Vectors using base R graphics: 
    wrap_elements(full = 
                    ~{
                      # plot coordinate axes
                      arrows3D(
                        x0 = 0, y0 = 0, z0 = 0,
                        x1 = 1, y1 = 0, z1 = 0,
                        col =  "black", 
                        theta = theta, phi = phi, expand = 0.8,
                        xlim = range(c(d_centered[1,], 0, 2, -2)),
                        ylim = range(c(d_centered[2,], 0, 2, -2)),
                        zlim = range(c(d_centered[3,], 0, 2, -2)),
                        scale = FALSE, box = FALSE, length = 0.1, lwd = .5 
                      )
                      segments3D(
                        x0 = max(d_centered[1,], 0), y0 = 0, z0 = 0,
                        x1 = min(d_centered[1,], 0), y1 = 0, z1 = 0,
                        col =  "black", lwd = 0.1, add = TRUE
                      )
                      arrows3D(
                        x0 = 0, y0 = 0, z0 = 0,
                        x1 = 0, y1 = 1, z1 = 0,
                        col =  "black", length = 0.1, lwd = .5, 
                        add = TRUE 
                      )
                      segments3D(
                        y0 = max(d_centered[2,], 0), x0 = 0, z0 = 0,
                        y1 = min(d_centered[2,], 0), x1 = 0, z1 = 0,
                        col =  "black", lwd = 0.1, add = TRUE
                      )
                      arrows3D(
                        x0 = 0, y0 = 0, z0 = 0,
                        x1 = 0, y1 = 0, z1 = 1,
                        col =  "black", length = 0.1, lwd = .5, 
                        add = TRUE
                      )
                      segments3D(
                        z0 = max(d_centered[3,], 0), x0 = 0, y0 = 0,
                        z1 = min(d_centered[3,], 0), x1 = 0, y1 = 0,
                        col =  "black", lwd = 0.1, add = TRUE
                      )
                      points3D(x = c(1.1,0,0), y = c(0,1.1,0), z = c(0,0,1.1), 
                               pch = c("i", "j", "k"), add = TRUE, 
                               col = "gray20", colvar = NULL)
                      # x-y/x-z/y-z plane
                      if(ij) rect3D (x0 = -2, y0 = -2, z0 = 0, 
                                     x1 = 2, y1 = 2, z1 = NULL, 
                                     col = scales::alpha("grey", .05), add = TRUE)
                      if(ik) rect3D (x0 = -2, y0 = 0, z0 = -2, 
                                     x1 = 2, y1 = NULL, z1 = 2, 
                                     col = scales::alpha("grey", .05), add = TRUE)
                      if(jk) rect3D (x0 = 0, y0 = -2, z0 = -2, 
                                     x1 = NULL, y1 = 2, z1 = 2, 
                                     col = scales::alpha("grey", .05), add = TRUE)
                      # Plot the centered vectors
                      arrows3D(
                        x0 = 0, y0 = 0, z0 = 0,
                        x1 = x_centered[1], y1 = x_centered[2], z1 = x_centered[3],
                        col = scales::alpha("red", .5), lwd = 2, length = 0.1, 
                        add = TRUE
                      )
                      arrows3D(
                        x0 = 0, y0 = 0, z0 = 0,
                        x1 = y_centered[1], y1 = y_centered[2], z1 = y_centered[3],
                        col = scales::alpha("blue", .5), lwd = 2, length = 0.1, 
                        add = TRUE
                      )
                      points3D(x = d_centered[1,]*1.1, y = d_centered[2,]*1.1, z = d_centered[3,]*1.1, 
                               pch = c("x", "y"), add = TRUE, 
                               col = c("red", "blue"), colvar = NULL)
                      
                      
                      # Add perpendicular lines to the planes
                      # For x vector
                      segments3D(
                        x0 = x_centered[1], y0 = x_centered[2], z0 = x_centered[3],
                        x1 = x_centered[1], y1 = x_centered[2], z1 = 0,
                        col = "red", lwd = .3, lty = 2, add = TRUE
                      )
                      segments3D(
                        x0 = x_centered[1], y0 = x_centered[2], z0 = x_centered[3],
                        x1 = x_centered[1], y1 = 0, z1 = x_centered[3],
                        col = "red", lwd = .3, lty = 2, add = TRUE
                      )
                      segments3D(
                        x0 = x_centered[1], y0 = x_centered[2], z0 = x_centered[3],
                        x1 = 0, y1 = x_centered[2], z1 = x_centered[3],
                        col = "red", lwd = .3, lty = 2, add = TRUE
                      )
                      
                      # For y vector
                      segments3D(
                        x0 = y_centered[1], y0 = y_centered[2], z0 = y_centered[3],
                        x1 = y_centered[1], y1 = y_centered[2], z1 = 0,
                        col = "blue", lwd = .3, lty = 2, add = TRUE
                      )
                      segments3D(
                        x0 = y_centered[1], y0 = y_centered[2], z0 = y_centered[3],
                        x1 = y_centered[1], y1 = 0, z1 = y_centered[3],
                        col = "blue", lwd = .3, lty = 2, add = TRUE
                      )
                      segments3D(
                        x0 = y_centered[1], y0 = y_centered[2], z0 = y_centered[3],
                        x1 = 0, y1 = y_centered[2], z1 = y_centered[3],
                        col = "blue", lwd = .3, lty = 2, add = TRUE
                      )
                    }, clip = TRUE
    )
  p
}
```


---

### 1. Positive Correlation

#### Data:

$$
\mathbf{x} = [1, 2, 3], \quad \mathbf{y} = [2, 4, 6]
$$

```{r, echo =FALSE}
data_matrix <- matrix(c(1, 2, 3, 2, 4, 6), nrow = 3, ncol = 2)
plot_vectors(data_matrix, jk = FALSE, theta = 30)
```

#### Step 1: Center the Data
Compute the means:

$$
\bar{x} = \frac{1 + 2 + 3}{3} = 2, \quad \bar{y} = \frac{2 + 4 + 6}{3} = 4
$$

Subtract the means to center the data:
$$
\mathbf{x_c} = [1 - 2, 2 - 2, 3 - 2] = [-1, 0, 1]
$$
$$
\mathbf{y_c} = [2 - 4, 4 - 4, 6 - 4] = [-2, 0, 2]
$$

#### Step 2: Compute the Inner Product
$$
\mathbf{x_c} \cdot \mathbf{y_c} = (-1)(-2) + (0)(0) + (1)(2) = 2 + 0 + 2 = 4
$$

#### Step 3: Compute the Norms
$$
\|\mathbf{x_c}\| = \sqrt{(-1)^2 + 0^2 + 1^2} = \sqrt{1 + 0 + 1} = \sqrt{2}
$$
$$
\|\mathbf{y_c}\| = \sqrt{(-2)^2 + 0^2 + 2^2} = \sqrt{4 + 0 + 4} = \sqrt{8} = 2\sqrt{2}
$$

#### Step 4: Compute the Correlation
$$
r_{XY} \frac{\mathbf{x_c} \cdot \mathbf{y_c}}{\|\mathbf{x_c}\| \|\mathbf{y_c}\|} = \frac{4}{\sqrt{2} \cdot 2\sqrt{2}} = \frac{4}{4} = 1
$$

#### Step 5: Compute the Angle
$$
\theta = \cos^{-1}(r) = \cos^{-1}(1) = 0^\circ
$$

#### Interpretation:
The vectors $\mathbf{x}$ and $\mathbf{y}$ are perfectly aligned ($\theta = 0^\circ$), indicating a **perfect positive correlation**:



---

### 2. Zero Correlation

#### Data:
$$
\mathbf{x} = [1, 2, 3], \quad \mathbf{y} = [1, -2, 1]
$$

```{r, echo = FALSE}
data_matrix <- matrix(c(1, 2, 3, 1, -2, 1), nrow = 3, ncol = 2)
plot_vectors(data_matrix, ik = FALSE, theta = 130)
```

#### Step 1: Center the Data
Compute the means:
$$
\bar{x} = 2, \quad \bar{y} = \frac{1 + (-2) + 1}{3} = 0
$$

Subtract the means:
$$
\mathbf{x_c} = [-1, 0, 1], \quad \mathbf{y_c} = [1, -2, 1]
$$

#### Step 2: Compute the Inner Product
$$
\mathbf{x_c} \cdot \mathbf{y_c} = (-1)(1) + (0)(-2) + (1)(1) = -1 + 0 + 1 = 0
$$

#### Step 3: Compute the Norms
$$
\|\mathbf{x_c}\| = \sqrt{2}, \quad \|\mathbf{y_c}\| = \sqrt{1^2 + (-2)^2 + 1^2} = \sqrt{6}
$$

#### Step 4: Compute the Correlation
$$
r_{XY} \frac{0}{\sqrt{2} \cdot \sqrt{6}} = 0
$$

#### Step 5: Compute the Angle
$$
\theta = \cos^{-1}(0) = 90^\circ
$$

#### Interpretation:
The vectors $\mathbf{x}$ and $\mathbf{y}$ are perpendicular ($\theta = 90^\circ$), indicating **zero correlation**.

---

### 3. Negative Correlation

#### Data:
$$
\mathbf{x} = [1, 2, 3], \quad \mathbf{y} = [3, 1, -1]
$$

```{r, echo = FALSE}
data_matrix <- matrix(c(1, 2, 3, 3, 1, -1), nrow = 3, ncol = 2)
plot_vectors(data_matrix, jk = FALSE)
```


#### Step 1: Center the Data
Compute the means:
$$
\bar{x} = 2, \quad \bar{y} = \frac{3 + 1 + (-1)}{3} = 1
$$

Subtract the means:
$$
\mathbf{x_c} = [-1, 0, 1], \quad \mathbf{y_c} = [2, 0, -2]
$$

#### Step 2: Compute the Inner Product
$$
\mathbf{x_c} \cdot \mathbf{y_c} = (-1)(2) + (0)(0) + (1)(-2) = -2 + 0 - 2 = -4
$$

#### Step 3: Compute the Norms
$$
\|\mathbf{x_c}\| = \sqrt{2}, \quad \|\mathbf{y_c}\| = \sqrt{2^2 + 0^2 + (-2)^2} = \sqrt{8} = 2\sqrt{2}
$$

#### Step 4: Compute the Correlation
$$
r_{XY} \frac{-4}{\sqrt{2} \cdot 2\sqrt{2}} = \frac{-4}{4} = -1
$$

#### Step 5: Compute the Angle
$$
\theta = \cos^{-1}(-1) = 180^\circ
$$

#### Interpretation:
The vectors $\mathbf{x}$ and $\mathbf{y}$ point in opposite directions ($\theta = 180^\circ$), indicating a **perfect negative correlation**.

---

### Summary of Results

| Correlation Type | $\mathbf{x}$| $\mathbf{y}$| $r_{XY}$| $\theta$|
|------------------|------------------|------------------|---------|--------------|
| Positive         | [1, 2, 3]        | [2, 4, 6]        | 1       | $0^\circ$|
| Zero             | [1, 2, 3]        | [1, -2, 1]       | 0       | $90^\circ$|
| Negative         | [1, 2, 3]        | [3, 1, -1]       | -1      | $180^\circ$|

These examples show how the correlation of data and the angle between vectors are deeply connected!

--------------------------------------------------------------

*(adapted from DeepSeek R1 output)*
